Bayesian RKHS-based methods in functional regression\support{Support information of the article.}

### 2022-02-28 paper/paper.tex
Update paper

%% Scalar product
\newcommand\dotprod[2]{\left\langle #1,\, #2 \right\rangle}

@@

\begin{abstract}
In this work we propose a Bayesian approach for functional linear or logistic  regression models, based on the theory of Reproducing Kernel Hilbert Spaces (RKHS's). These new models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional alternative to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subset of the RKHS generated by the underlying covariance function. By imposing a suitable prior distribution on such RKHS, we can perform data-driven inference via standard Bayes methodology. The posterior distribution can be approximated from Markov Chain Monte Carlo (MCMC) methods. Several estimators  derived from this posterior distribution turn out to be competitive against other usual  alternatives in both simulated examples and real datasets.
\end{abstract}

@@

The problem of inferring a scalar response \(Y\) from a functional covariate \(X=X(t)\) is one that has gained traction over the last few decades, as more and more data is being generated with an ever-increasing level of granularity in the measurements. While in principle we could simply treat the functional data as a discretized vector in a very high dimension, there are often many advantages in taking into account the functional nature of the data, ranging from modeling the possibly high correlation among points that are close in the domain, to extracting information that may be hidden in the derivatives of the function in question. As a consequence, numerous proposals have arisen on how to suitably treat functional data, all of them encompassed under the term Functional Data Analysis (FDA), which essentially explores statistical techniques to process, model and make inference on continuous data. A recent survey on such techniques and methods is \citet{cuevas2014partial}, while a more detailed exposition of the theory and applications can be found for example in \citet{hsing2015theoretical} or \citet{berrendero2019rkhs}.

In this work we are concerned with (supervised) functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges that emerge as a result of working in infinite dimensions. To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some hidden dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\). We will assume the existence of a ``labeled'' data set \(\mathcal D_n =\{(X_i(t), Y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), and our aim will be to accurately predict the response corresponding to unlabeled samples from \(X\).

The most common linear regression model is the classical functional linear regression model, originally introduced in the first edition of the book by~\citet{ramsay2005functional} \incomment{fueron realmente los primeros?}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^n\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \dotprod{X}{\beta} + \epsilon = \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
where \(\epsilon\) is an independent error term with certain simplifying assumptions. In this case, the main parameter \(\beta=\beta(t)\) is a member of the infinte-dimensional space \(L^2[0, 1]\). Aside from the fact that functional (i.e. infinite-dimensional) optimization is a hard problem in and of itself \incomment{cita?}, there are a number of more subtle drawbacks that arise in this \(L^2\)-framework. For instance, \(L^2[0, 1]\) is a big space that contains many non-smooth or ill-behaved functions, but incidentally the model given by~\eqref{eq:l2-linear-model} does not include a finite-dimensional model based on a linear combination of projections of \(X\) as a particular case (see \citet{berrendero2019rkhs}). Moreover, the non-invertibility of the covariance operator (defined in Section~\ref{sec:rkhs}), which plays the role of the covariance matrix in the infinite case, invalidates the usual least-squares theory.

A similar \(L^2\)-model can be derived for the binary classification problem via the logistic function:
\begin{equation}\label{l2-logistic-model}
  \mathbb P(Y \mid X(t)) = \frac{1}{}
\end{equation}

It turns out that a natural alternative to the \(L^2\) model is the RKHS model..... Not only simpler, but also smoother and more manageable functions. Sustituir producto escalar por el de \(H(K)\). This model has already been proposed in \incomment{cita Berrendero et al. 2019}, and in \incomment{cita Torrecilla 2017} for the case of functional classification.

Bayesian RKHS-based variable selection (cite Berrendero et als.).

@@

As we can see, the covariance function (sometimes referred to as the \textit{kernel}) plays a crucial role in characterizing the RKHS \incomment{¿Es necesario destacar alguna suposición/propiedad extra sobre \(K\)? Por ejemplo \(\int\int K^2 <\infty\) o que es simétrica y (semi-) definida positiva}. A compact operator closely related to this covariance function is the so-called covariance operator, i.e.: \incomment{No sé muy bien qué calificativos ponerle: integral Hilbert-Schmidt operator, compact, self-adjoint, bounded, linear, ...}
\[
\mathcal Kf(\cdot) = \int_0^1 K(s, \cdot)f(s)\, ds, \quad f \in L^2[0, 1].
\]
This operator is especially relevant due to the \textit{Karhunen-Loève} theorem \incomment{cita}, which asserts that a centered process \(X\) can be decomposed as
\begin{equation}\label{eq:karhunen-loeve}
X(t) = \sum_{j=1}^\infty \xi_j \phi_j(t),
\end{equation}
where \(\phi_j\) are the (orthonormal) eigenfunctions of its covariance operator \(\mathcal K\) and \(\xi_j = \int X\phi_j\) are uncorrelated zero-mean random variables with variance equal to \(\lambda_j\), the eigenvalue associated with \(\phi_j\). This expansion is the main justification of many FDA techniques, such as functional principal component analysis (FPCA) (see for example \incomment{cita}).

@@

\begin{enumerate}
  \item Via \textit{Loeve's isometry}, one can establish a congruence between \(\mathcal H(K)\) and the linear span of the centered process, \(\mathcal L(X)\), in the space of all random variables with finite second moment, \(L^2(\Omega)\) \incomment{cita}. This isometry is essentially the completion of the correspondence
  \[
  \sum_{i=1}^p a_i (X(t_i) - m(t_i)) \longleftrightarrow \sum_{i=1}^p a_i K(t_i, \cdot),
\]
and can be formally defined as (see \incomment{cita})
\begin{equation}\label{eq:loeves-isometry}
  \Psi_X(U)(t) = \E[U(X(t) - m(t))], \quad U \in \mathcal L(X).
\end{equation}
  \item The space \(\mathcal H(K)\) can also be defined as the image of the square root of the covariance operator, i.e.:
  \begin{equation}\label{eq:rkhs-square-root}
  \mathcal H(K) = \mathcal K^{1/2}(L^2[0, 1]).
\end{equation}
In this case, the inner product can be expressed as
\[
\dotprod{f}{g}_K = \dotprod{\mathcal K^{1/2}(f)}{\mathcal K^{1/2}(g)}.
\]

@@

\section{Bayesian methodology for RKHS-based functional regression models}

@@

  - Parametric approach with \(\beta \in H_0(K)\) (further constrain the parameter to this subset of the RKHS, which is dense by construction). Then we can impose a prior on the finite-dimensional parameter space.
\end{comment}

@@

@@

\subsection{Functional linear regression}
 The model remains linear because of the correspondence between the RKHS and the linear span of the process (Loève’s isometry). The RKHS model includes both L2 model and finite model. Cite Berrendero et al 2020.

@@

In this case the model/functional equation in $L^2$ is....... and the same problems apply, plus the MLE existence problem. An RKHS aproach is given by:.......

 and this solves the problems (see Bueno-Larraz 2021)

@@

\begin{comment}
  Not only MSE/Acc but also credible intervals, points-of-impact model --> variable selection included.
\end{comment}

>>>>

%% Scalar product
\newcommand\dotprod[2]{\left\langle #1, #2 \right\rangle}

@@

\begin{abstract}
In this work we propose a Bayesian approach for functional linear or logistic  regression models, based on the theory of Reproducing Kernel Hilbert Spaces (RKHS's). These new models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional alternative to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subset of the RKHS generated by the underlying covariance function. By imposing a suitable prior distribution on such RKHS, we can perform data-driven inference via standard Bayes methodology. The posterior distribution can be approximated from Markov Chain Monte Carlo (MCMC) methods. Several estimators derived from this posterior distribution turn out to be competitive against other usual  alternatives in both simulated examples and real datasets, including a Bayesian-motivated variable selection method.
\end{abstract}

@@

The problem of inferring a scalar response from a functional covariate is one that has gained traction over the last few decades, as more and more data is being generated with an ever-increasing level of granularity in the measurements. While in principle we could simply treat the functional data as a discretized vector in a very high dimension, there are often many advantages in taking into account the functional nature of the data, ranging from modeling the possibly high correlation among points that are close in the domain, to extracting information that may be hidden in the derivatives of the function in question. As a consequence, numerous proposals have arisen on how to suitably treat functional data, all of them encompassed under the term Functional Data Analysis (FDA), which essentially explores statistical techniques to process, model and make inference on continuous data. A recent survey on such techniques and methods is \citet{cuevas2014partial}, while a more detailed exposition of the theory and applications can be found for example in \citet{hsing2015theoretical} and \citet{horváth2012inference}.

In this work we are concerned with (supervised) functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges that emerge as a result of working in infinite dimensions. To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some hidden dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\). We will assume the existence of a ``labeled'' data set \(\mathcal D_n =\{(X_i, Y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), and our aim will be to accurately predict the response corresponding to unlabeled samples from \(X\).

The most common linear regression model is the classical functional linear regression model, originally introduced in the first edition of the book by~\citet{ramsay2005functional} \incomment{fueron realmente los primeros?}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^n\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \dotprod{X}{\beta} + \epsilon = \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
\incomment{mejor poner el producto con el subíndice \(\langle \cdot , \cdot \rangle_2\)?}~where \(\epsilon\) is an independent error term with certain simplifying assumptions. In this case, the main parameter \(\beta=\beta(t)\) is a member of the infinte-dimensional space \(L^2[0, 1]\). Aside from the fact that functional (i.e. infinite-dimensional) optimization is a hard problem in and of itself \incomment{cita?}, there are a number of more subtle drawbacks that arise in this \(L^2\)-framework. For instance, \(L^2[0, 1]\) is a big space that contains many non-smooth or ill-behaved functions, but incidentally the model given by~\eqref{eq:l2-linear-model} does not include a finite-dimensional model based on a linear combination of projections of \(X\) as a particular case (see \citet{berrendero2019rkhs}). Moreover, the non-invertibility of the covariance operator (defined in Section~\ref{sec:rkhs}), which plays the role of the covariance matrix in the infinite case, invalidates the usual least squares theory.

A similar \(L^2\)-based logistic regression model can be derived for the binary classification problem via the logistic function:
\begin{equation}\label{l2-logistic-model}
  \mathbb P(Y=1 \mid X=x) = \frac{1}{1 + \exp\{-\beta_0 - \dotprod{\beta}{x}\}},
\end{equation}
where \(\beta_0 \in \R\) and \(\beta \in L^2[0, 1]\). In this case, the most common way of estimating the slope function \(\beta(t)\) is via its Maximum Likelihood Estimator (MLE). However, not only do the same complications as in the linear regression model apply in this situation, but also there is the additional problem that in functional settings the MLE does not exist with probability one under fairly general conditions (see~\citet[p.~10]{buenolarraz2021functional}).

It turns out that a natural alternative to the \(L^2\) model is the so-called Reproducing Kernel Hilbert Space (RKHS) model, which instead assumes the unknown parameter to be a member of the RKHS associated with the covariance function of the process \(X\), making use of the scalar product of that space. As we will show later on, not only is this model simpler and arguably easier to interpret, but it also constrains the parameter space to smoother and more manageable functions. This approach has already been proposed in \citet{berrendero2019rkhs} in the functional linear regression setting, and in \citet{berrendero2018use} for the case of functional classification.

Having provided a general perspective on the problem we are trying to solve, our main contribution is a novel Bayesian approach to linear and logistic regression with functional data, building on the newly introduced RKHS models for these tasks, and deviating from the usual methods of basis expansion and finite-dimensional least squares for parameter estimation. In essence, we suggest to impose a prior distribution on the RKHS associated with the functional regressor, and perform inference on this parametric space via standard Bayesian methodology. Luckily, the posterior distribution can be numerically approximated with Markov Chain Monte Carlo (MCMC) methods (see Section \ref{sec:methodology}), so that a complete end-to-end estimator can be derived. Moreover, the fact that we have a posterior distribution to work with allows us to perform diverse kinds of analyses to validate and assess the model. Similar ideas were presented in \citet{grollemund2019bayesian}, but the authors did not consider an RKHS-based model.

Another set of techniques widely studied in this context are variable selection methods, which aim to select the marginals \(\{X(t_i)\}\) of the process that better summarize it according to some optimality criterion. As it happens, some variable selection methods have already been proposed, both in the RKHS framework and in more general settings (see for example \citet{berrendero2019rkhs}, \citet{berrendero2016mrmr} or \citet{berrendero2016variable}), but in general they have their own dedicated algorithms and procedures. Given the nature of our proposed Bayesian model, we can easily isolate the marginal posterior distribution corresponding to a finite set of points of impact \(\{t_i\}\), thus providing a variable selection method along with the other estimators that naturally arise within our model. In this way, we can not only make predictions about our data, but also evaluate exactly which marginals of the process contain the most information.

@@

As we can see, the covariance function (sometimes referred to as the \textit{kernel}) plays a crucial role in characterizing the RKHS \incomment{¿Es necesario destacar alguna suposición/propiedad extra sobre \(K\)? Por ejemplo \(\int\int K^2 <\infty\) o que es simétrico y (semi-) definido positivo}. A linear operator closely related to this covariance function is the so-called covariance operator, i.e.: \incomment{No sé muy bien qué calificativos ponerle: integral Hilbert-Schmidt operator, compact, self-adjoint, bounded, linear functional, ...}
\[
\mathcal Kf(\cdot) = \int_0^1 K(s, \cdot)f(s)\, ds, \quad f \in L^2[0, 1].
\]
This operator is especially relevant due to the \textit{Karhunen-Loève} theorem \incomment{cita}, which asserts that a centered process \(X\) can be decomposed as
\begin{equation}\label{eq:karhunen-loeve}
X(t) = \sum_{j=1}^\infty \xi_j \phi_j(t),
\end{equation}
where \(\phi_j\) are the (orthonormal) eigenfunctions of its covariance operator \(\mathcal K\) and \(\xi_j = \int X\phi_j\) are uncorrelated zero-mean random variables with variance equal to \(\lambda_j\), the eigenvalue associated with \(\phi_j\). This expansion is the main justification of many FDA techniques, such as functional principal component analysis (FPCA) (see for example \incomment{cita}).

@@

\begin{enumerate}
  \item Via \textit{Loeve's isometry}, one can establish a congruence between \(\mathcal H(K)\) and the linear span of the centered process, \(\mathcal L(X)\), in the space of all random variables with finite second moment, \(L^2(\Omega)\) \incomment{cita}. This isometry is essentially the completion of the correspondence
  \[
  \sum_{i=1}^p a_i (X(t_i) - m(t_i)) \longleftrightarrow \sum_{i=1}^p a_i K(t_i, \cdot),
\]
and can be formally defined as (see \incomment{cita})
\begin{equation}\label{eq:loeves-isometry}
  \Psi_X(U)(t) = \E[U(X(t) - m(t))], \quad U \in \mathcal L(X).
\end{equation}
  \item The space \(\mathcal H(K)\) can also be identified with the image of the square root of the covariance operator, i.e.:
  \begin{equation}\label{eq:rkhs-square-root}
  \mathcal H(K) = \mathcal K^{1/2}(L^2[0, 1]).
\end{equation}
In this case, the inner product can be expressed as
\[
\dotprod{f}{g}_K = \dotprod{\mathcal K^{1/2}(f)}{\mathcal K^{1/2}(g)}.
\]

@@

\section{Bayesian methodology for RKHS-based functional regression models}\label{sec:methodology}

@@

  - Parametric approach with \(\beta \in H_0(K)\) (further constrain the parameter to this subset of the RKHS, which is dense by construction). Then we can impose a prior on the finite-dimensional parameter space.

  - Substitute scalar product of L2 for that of H(K).
\end{comment}

@@

\begin{comment}
Bayesian variable selection: select marginal and then apply multivariate methods.
\end{comment}

@@

\subsection{Functional linear regression}
 The model remains linear because of the correspondence between the RKHS and the linear span of the process (Loève’s isometry). The RKHS model includes both L2 model and finite model (most problems solved). Cite Berrendero et al 2020.

@@

An RKHS aproach is given by:.......

As well as the other problems, The problem of MLE are solved, and to avoid the problem in finite dimension, we could use the strategy in Bueno Larraz 2021 (Firth + 2 phase algorithm for t and beta).

@@

\begin{comment}
  Not only MSE/Acc but also credible intervals, points-of-impact model --> variable selection included.

  - Comparar con Grollemund et al. ?
\end{comment}
